---
title: "Road casualties in GM"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    theme: readable
    logo: trafford_council_logo_black_on_white_100px.png
    css: styles.css
    source_code: "https://github.com/trafforddatalab/road_casualties"
    includes:
      after_body: footer.html
runtime: shiny
---


```{r setup, include=FALSE}
library(tidyverse) ; library(flexdashboard) ; library(shiny) ;  library(sf) ; library(leaflet) ;  library(scales) ; library(shinyWidgets) ; library(plotly) ; library(zoo)

    
  df <- read_csv("https://www.trafforddatalab.io/open_data/road_casualties/STATS19_road_casualties_2010-2021.csv") %>% 
    mutate(AREFNO = as.character(AREFNO)) %>% 
    filter(!year %in% c(2019,2020,2021)) %>%
    bind_rows(read_csv("https://www.trafforddatalab.io/open_data/road_casualties/STATS19_road_casualties_2019-Jun2024.csv") ) %>%
  mutate(text = str_c("On ", format.Date(date, "%A %d %B %Y"), " at ", format(strptime(output_time,"%H:%M:%S"),'%H:%M'),
                      " a ", tolower(collision_severity), " collision occured involving ", 
                      number_vehicles, ifelse(number_vehicles == 1, " vehicle and ", " vehicles and "), number_casualties, 
                      ifelse(number_casualties == 1, " casualty.", " casualties.")))

boundaries <- st_read("data/gm_local_authority_generalised.geojson")


```

```{r reactive}
casualties <- reactive({
  
  if((is.null(input$borough) & is.null(input$mode) & is.null(input$severity))){filter(df, date >= input$start_month &  date <= as.yearmon(paste(input$end_month)) %>% as.Date(ym, frac = 1))}else 
  if((is.null(input$borough) & is.null(input$mode))){filter(df, date >= input$start_month &  date <= as.yearmon(paste(input$end_month)) %>% as.Date(ym, frac = 1) &
             casualty_severity %in% input$severity)}else 
  if((is.null(input$borough) & is.null(input$severity))){filter(df, date >= input$start_month &  date <= as.yearmon(paste(input$end_month)) %>% as.Date(ym, frac = 1) &
              mode %in% input$mode)}else 
  if((is.null(input$mode) & is.null(input$severity))){filter(df, date >= input$start_month &  date <= as.yearmon(paste(input$end_month)) %>% as.Date(ym, frac = 1) &
              area_name %in% input$borough)}else 
  if((is.null(input$borough))){filter(df, date >= input$start_month &  date <= as.yearmon(paste(input$end_month)) %>% as.Date(ym, frac = 1) &
              mode %in% input$mode &
              casualty_severity %in% input$severity)}else 
  if((is.null(input$mode))){filter(df, date >= input$start_month &  date <= as.yearmon(paste(input$end_month)) %>% as.Date(ym, frac = 1) &
            area_name %in% input$borough &
            casualty_severity %in% input$severity)}else 
  if((is.null(input$severity))){filter(df, date >= input$start_month &  date <= as.yearmon(paste(input$end_month)) %>% as.Date(ym, frac = 1) &
            area_name %in% input$borough &
            mode %in% input$mode)}else {
    filter(df, date >= input$start_month &  date <= as.yearmon(paste(input$end_month)) %>% as.Date(ym, frac = 1) &
             area_name %in% input$borough &
             mode %in% input$mode &
             casualty_severity %in% input$severity)
  }
})

collisions <- reactive({distinct(casualties(), AREFNO, .keep_all = TRUE)})


```

Filter {.sidebar data-width=200}
=======================================================================

```{r}

airDatepickerInput("start_month",
                     label = "Start month",
                     value = "2024-01-01",
                     maxDate = "2024-06-30",
                     minDate = "2010-01-01",
                     view = "months", 
                     minView = "months", 
                     dateFormat = "yyyy-MM")

airDatepickerInput("end_month",
                     label = "End month",
                     value = "2024-06-30",
                     maxDate = "2024-06-30",
                     minDate = "2010-01-01",
                     view = "months", 
                     minView = "months", 
                     dateFormat = "yyyy-MM")

selectInput("borough", 
            label = "Local Authority", 
            choices = c("All" = "", unique(df$area_name)),
            multiple = TRUE)

selectInput("mode", 
            label = "Mode of travel",
            choices = c("All" = "", unique(df$mode)),
            multiple = TRUE)

selectInput("severity", 
            label = "Casualty severity",
            choices = c("All" = "", unique(df$casualty_severity)),
            multiple = TRUE)


```

Map
=====================================  


### Road Accidents by most severe casualty (Accidents could have one or more casualties with different severity each) {data-width=700}
```{r}
pal <- colorFactor(c("#b10026", "#fd8d3c", "#ffeda0"), domain = c("Fatal", "Serious", "Slight"), ordered = TRUE)

output$map1 <- renderLeaflet({
  
  validate(need(nrow(collisions())!=0, "There are no matches in the dataset. Choose other categories"))
  
  leaflet(data = collisions()) %>%
    addTiles(urlTemplate = "https://cartodb-basemaps-{s}.global.ssl.fastly.net/light_all/{z}/{x}/{y}.png",
               attribution = '&copy; <a href="http://www.openstreetmap.org/copyright">OpenStreetMap</a>, <a href="http://cartodb.com/attributions">CartoDB</a> | <a href="https://www.ons.gov.uk/methodology/geography/licences">Contains OS data © Crown copyright and database right (2019)</a>',
               group = "CartoDB",
               options = providerTileOptions(minZoom = 10, maxZoom = 16)) %>%
    addPolygons(data = boundaries, fill = F, color = "#636363", weight = 1.5) %>% 
    addCircleMarkers(~lon, ~lat, 
                         color = "#636363", stroke = TRUE, weight = 1, 
                         fillColor = ~pal(collision_severity), fillOpacity = 0.8, 
                         radius = 5, popup = ~text) %>% 
    fitBounds(lng1 = max(collisions()$lon), lat1 = max(collisions()$lat),
              lng2 = min(collisions()$lon), lat2 = min(collisions()$lat)) %>% 
    addLegend(position = "bottomright", colors = c("#b10026", "#fd8d3c", "#ffeda0"),
              labels = c("Fatal", "Serious", "Slight"), opacity = 1, title = NULL)
  })

leafletOutput("map1")

observeEvent(input$active_tab, {
  if (input$active_tab == 0) {

  leafletProxy("map1", session) 

  }
  })

```   

### By Local authority {data-width=340}

```{r}
renderTable({
  collisions() %>%
    group_by(area_name) %>%
    summarise(collisions = n(), casualties = sum(number_casualties)) %>%
    arrange(desc(collisions)) %>%
    select("Local Authority" = area_name, "Accidents" = collisions, "Casualties" = casualties) %>%
    as.data.frame()
}, digits = 0)


```

Charts
=====================================

Row
-----------------------------------------------------------------------

### Trend
```{r}
output$timeline <- renderPlot({
  
empty <-  tibble(date= seq(input$start_month, as.yearmon(paste(input$end_month)) %>% as.Date(ym, frac = 1), by = "1 months"))

temp <- casualties() %>%
  group_by(date = lubridate::floor_date(date, "month")) %>%
    summarize(value = n())

df <- left_join(empty, temp, by = "date") %>%
  replace_na(list(value = 0))
  
ggplot(df, aes(x = date, y = value)) +
  geom_line(size = 1.5, color = "#0A3182") +
  geom_hline(yintercept = 0, size = 1, colour = "#333333") +
 scale_y_continuous(expand = c(0.005, 0.005),
                     breaks = function(x) unique(floor(pretty(seq(0, (max(x) + 1) * 1.1)))), labels = comma) +
  #scale_x_date(date_labels = "%b %y") +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Casualties", 
       caption = paste0(format(input$start_month,"%h-%Y"), " to ", format(as.yearmon(paste(input$end_month)) %>% as.Date(ym, frac = 1),"%h-%Y")," Source: TfGM")
       ) +
  theme_minimal(base_size = 10) +
  theme(plot.margin = unit(c(1,1,1,1), "cm"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.caption = element_text(color = "grey50"),
        axis.title.y = element_text(hjust = 1),
        legend.position = "bottom")

})

plotOutput('timeline')

```

### By Month
```{r}
output$months2 <- renderPlot({

empty <- tibble(casualty_severity = c(rep("Slight", 12), rep("Serious", 12), rep("Fatal", 12)),
                month = rep(month.abb, 3)) %>% 
  mutate(id = paste(casualty_severity, month, sep = '_'))

temp <- casualties() %>%
  group_by(casualty_severity, month) %>%
  summarise(value = n()) %>% 
  ungroup() %>% 
  mutate(id = paste(casualty_severity, month, sep = '_')) %>%
  select(-casualty_severity, -month)

df <- left_join(empty, temp, by = "id") %>% 
  mutate(month = factor(month, levels = month.abb))
  
ggplot(df, aes(x = month, y = value)) +
  geom_col(aes(fill = factor(casualty_severity)), size = 1.5) +
  geom_hline(yintercept = 0, size = 1, colour = "#333333") +
  scale_fill_manual(values = 
                      c("Fatal" = "#b10026",
                        "Serious" = "#fd8d3c",
                        "Slight" = "#ffeda0"),
                    drop = FALSE,
                    guide = guide_legend(keyheight = unit(5, units = "mm"), 
                                            keywidth=unit(5, units = "mm")) ) +
 scale_y_continuous(expand = c(0.005, 0.005), 
                     breaks = function(x) unique(floor(pretty(seq(0, (max(x) + 1) * 1.1)))), labels = comma) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Casualties by month and severity", 
      caption = paste0(format(input$start_month,"%h-%Y"), " to ", format(input$end_month,"%h-%Y")," Source: TfGM")) +
  theme_minimal(base_size = 10) +
  theme(plot.margin = unit(c(1,1,1,1), "cm"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.caption = element_text(color = "grey50"),
        axis.title.y = element_text(hjust = 1),
        legend.position = "right")


})


plotOutput('months2')

```

Row
-----------------------------------------------------------------------

### By Hour

```{r}
output$hours <- renderPlot({

empty <- tibble(day = c(rep("Mon", 24), rep("Tue", 24), rep("Wed", 24),
               rep("Thu", 24), rep("Fri", 24), rep("Sat", 24),
               rep("Sun", 24)),
       hour = rep(seq(0, 23, by = 1), 7)) %>% 
  mutate(id = paste(day, hour, sep = '_'))
  
temp <- casualties() %>%
  group_by(day, hour) %>%
  summarise(value = n()) %>% 
  ungroup() %>% 
  mutate(id = paste(day, hour, sep = '_')) %>%
  select(-day, -hour)

df <- left_join(empty, temp, by = "id") %>% 
  mutate(day = factor(day, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"), ordered = TRUE),
         hour = factor(hour))

ggplot(df, aes(x = factor(hour), y = fct_rev(day))) +
  geom_tile(aes(fill = value), colour = "white", size = 1) + 
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "#ffffff",
                       limits = c(1, NA),
                       breaks = function(x) unique(floor(pretty(seq(0, (max(x) + 1) * 1.1)))),
                       guide = guide_colorbar(direction = "vertical",
                                              #barheight = unit(2, units = "mm"), 
                                              barwidth = unit(3, units = "mm"),
                                              label.hjust = 0.5)) +
  scale_x_discrete(breaks = 0:23, labels = 0:23, drop = FALSE) + 
  scale_y_discrete(expand = c(0,0)) +
  labs(x = NULL, y = NULL, fill = NULL, 
       title = "Casualties by week day and hour", 
              caption = paste0(format(input$start_month,"%h-%Y"), " to ", format(input$end_month,"%h-%Y")," Source: TfGM")) +
  theme_minimal(base_size = 10) +
  theme(plot.margin = unit(c(1,1,1,1), "cm"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.caption = element_text(color = "grey50"),
        legend.position = "right") +
  coord_equal(ratio = 1)

})

plotOutput('hours')
```

### Demographics

```{r}
output$demographics <- renderPlot({
  
empty <- tibble(sex = c(rep("Female", 11), rep("Male", 11)),
                ageband = rep(c("0-5","6-10","11-15","16-20","21-25","26-35","36-45","46-55","56-65",
                                "66-75","Over 75"), 2)) %>% 
  mutate(id = paste(sex, ageband, sep = '_'))

temp <- casualties() %>%
     group_by(sex, ageband) %>%
  summarise(value = n()) %>%
  ungroup() %>% 
  mutate(id = paste(sex, ageband, sep = '_')) %>%
  select(-sex, -ageband)

df <- left_join(empty, temp, by = "id") %>% 
  mutate(sex = factor(sex),
         ageband = fct_relevel(ageband, "0-5", "6-10"))
  
  ggplot(df, aes(x = ageband, y = value, group = sex, fill = sex)) + 
  geom_col(position = "dodge", alpha = 0.8) +
      geom_hline(yintercept = 0, size = 1, colour = "#333333") +
    scale_y_continuous(expand = c(0.005, 0.005), 
                     breaks = function(x) unique(floor(pretty(seq(0, (max(x) + 1) * 1.1)))), labels = comma) +
  scale_fill_manual(values = 
                      c("Female" = "#194a81",
                        "Male" = "#1d85a5"),
                    guide = guide_legend(keyheight = unit(3, units = "mm"), 
                                            keywidth=unit(5, units = "mm")) ) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Casualties by ageband and sex", 
      caption = paste0(format(input$start_month,"%h-%Y"), " to ", format(input$end_month,"%h-%Y")," Source: TfGM")) +
  theme_minimal(base_size = 10) +
  theme(plot.margin = unit(c(1,1,1,1), "cm"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.caption = element_text(color = "grey50"),
        axis.title.y = element_text(hjust = 1),
        legend.position = "bottom")
  
})

plotOutput('demographics')

```

About
=====================================  

This application allows users to explore road collisions involving personal injury that were reported in Greater Manchester between 2010 and June 2024
. The data derive from <a href="https://data.gov.uk/dataset/25170a92-0736-4090-baea-bf6add82d118/gm-road-casualty-accidents-full-stats19-data">Transport for Greater Manchester</a> under an <a href="http://odata.tfgm.com/opendata/downloads/Licences/FreeToUseData_SubLicence_G_v12.pdf">OS "Free to Use Data" Licence</a> and from <a href="https://data.gov.uk/dataset/25170a92-0736-4090-baea-bf6add82d118/gm-road-casualty-accidents-full-stats19-data">Department from Transport</a> under an <a href="https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/">Open Government License</a>.

<script>
$(document).ready(function(){
  $("body").on("shown.bs.tab", "a[data-toggle='tab']", function(e) {
    Shiny.setInputValue("active_tab", $(e.target).parent().index(), {priority: "event"});
  });
});
</script>


